---
title: "Loading Data into a Data Frame"
author: "Gabriella Martinez"
date: "2/6/2021"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
## Super Bowl Advertisements
### Overview
With the super bowl around the corner, the data set "superbowl -ads" from fivethirtyeight <https://projects.fivethirtyeight.com/super-bowl-ads/> delves into the top ten brands with the most advertisements run on Super Bowl Sundays from 2000 to 2020. The data for "superbowl-ads" originates from <http://www.superbowl-ads.com>. The data set includes a link to the YouTube video for the advertisement and seven straight forward labeled attributes used to describe the advertisement which are: funny, show_product_quickly, patriotic, celebrity, danger, animals, and use_sex. 

```{r load_libraries, include=FALSE}
library(RCurl)
library(colorspace)
library(tidyverse)
library(ggplot2)
```
```{r fetch_plot, echo=FALSE}
x <- url("https://raw.githubusercontent.com/fivethirtyeight/superbowl-ads/main/superbowl-ads.csv")
superbowl_ads <- read.csv(x)

funny_plot <- ggplot(superbowl_ads, aes(x = funny)) +
        geom_bar(aes(y = (..count..)/sum(..count..))) +
        xlab("Funny Commercials")+
        scale_y_continuous(labels = scales::percent, name = "Proportion") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(funny_plot + ggtitle("Proportion of Funny Super Bowl Commercials 2000-2020 "))


```

### Findings and Recommendations  
Since the attributes funny, show_product_quickly, patriotic, celebrity, danger, animals, and use_sex are qualitative in nature, it would be a good idea to obtain the amount of revenue generated as a result of the advertisements run on Super Bowl Sunday and see which types and combinations of qualitative attributes produce the most revenue for each brand.  

#### Super Bowl Summary Data
```{r additional_data, echo=FALSE}
summary(superbowl_ads)
```

Breakdown of 'Funny' Attribute
```{r additional_dataa, echo=FALSE}
table(superbowl_ads$funny)
```

