---
title: "R Character Manipulation and Date Processing"
author: "Gabriella Martinez"
date: "2/17/2021"
output:
  html_document: default
  pdf_document: default
---

#### 1. Provide code that identifies the majors that contain either "DATA" or "STATISTICS" from fivethirtyeight.com’s College Majors dataset.

```{r load_libraries, include=FALSE}
library(RCurl)
library(colorspace)
library(tidyverse)
library(ggplot2)
library(dbplyr)
```
```{r load_data}
x <- url("https://raw.githubusercontent.com/fivethirtyeight/data/master/college-majors/majors-list.csv")
college_majors <- read.csv(x)

majors <- as.matrix.data.frame(college_majors)

grep(pattern = 'STATISTICS', majors, value = TRUE, ignore.case = TRUE)
grep(pattern = 'DATA', majors, value = TRUE, ignore.case = TRUE)

```

#### 2. Write code that transforms the data below.  
\n  

[1] "bell pepper"  "bilberry"     "blackberry"   "blood orange"

[5] "blueberry"    "cantaloupe"   "chili pepper" "cloudberry"  

[9] "elderberry"   "lime"         "lychee"       "mulberry"    

[13] "olive"       "salal berry"

```{r}
#making the data into a list
x <- list('[1]', "bell pepper",  "bilberry",     "blackberry",   "blood orange",

'[5]', "blueberry",    "cantaloupe",   "chili pepper", "cloudberry",  

'[9]', "elderberry",   "lime",         "lychee",       "mulberry",    

'[13]', "olive",       "salal berry", " ", " ")

#makes the list into a matrix https://www.tutorialspoint.com/how-to-convert-a-list-to-matrix-in-r
matrix_x <- matrix(unlist(x), ncol = 5, nrow=4, byrow=TRUE)

#removes the first column of the matrix
matrix_x <- matrix_x[, -1]

#transposes the matrix to then covert into vector that prints the items in order
vector_x <- as.vector(t(matrix_x))

#removes last two " " from the vector
head(vector_x, -2)


```


#### 3. Describe, in words, what these expressions will match:
```{r list}
fruits_veg <- c("apple", "banana", "pear", "papaya", "coconut", "cucumber", "jujube", "pepper")
```

(.)\1\1  
The (.) captures the first character while \1 and \1 don't perform a task because proper backreferencing command wasn't executed.
```{r}
str_view(fruits_veg, "(.)\1\1", match = TRUE)
```
(.)(.)\\2\\1  
The first (.) captures a single character, the second (.) captures a different single character, \\2 backreferences the second character, backreferences the first character.
```{r}
str_view(fruits_veg, "(.)(.)\\2\\1", match = TRUE)
```
(..)\1  
The (..) captures two characters together as a group, but command will not match anything because proper exiting and backreferencing was not executed.
```{r}
str_view(fruits_veg, "(..)\1", match = TRUE)
```

(.).\\1.\\1  
The (.) captures one character, . recognizes another character distinct from the first, \\1 backreferences the first character, . allows for any character, \\1 backreferences the first character again.  
```{r}
str_view(fruits_veg, "(.).\\1.\\1", match = TRUE)
```

(.)(.)(.).*\\3\\2\\1  
The (.)(.)(.) captures three distinct characters, .* captures any character inlcuding one of those already recorded, \\3\\2\\1 reverses the order of the first three characters identified.
```{r}
name <- c("habibah")
str_view(name, "(.)(.)(.).*\\3\\2\\1", match = TRUE)
```
 

#### 4. Construct regular expressions to match words that:

Start and end with the same character.

```{r}
names <- c("itati", "anabella", "ariella")
words_1 <-c("dated", "going", "devised")
str_view(names, "^(.).*\\1$", match = TRUE)
str_view(words_1, "^(.).*\\1$", match = TRUE)
```
Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)
```{r}
words_2 <- c("church", "papaya", "banana", "curcurma")
str_view(words_2, "(..).*\\1", match = TRUE)
```
  
Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)

```{r}
words_3 <- c("havana", "banana","papaya", "tallahasee","kissimee")
str_view(words_3, "(.).*\\1.*\\1", match = TRUE)
```

Just for fun:
```{r}
words_4 <- c("mississippi","tallahasee","kissimee", "abcttxyzllpoik")
str_view(words_4, "(..*)*\\1", match = TRUE)
```