---
title: "Web APIs"
author: "Gabriella Martinez"
date: "4/7/2021"
output: 
    html_document:
        toc: true
        toc_float: true
        number_sections: false
        theme: sandstone
        highlight: tango
        font-family: "Arial"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# "default", "cerulean", "journal", "flatly", "darkly", "readable", "spacelab", "united", "cosmo", "lumen", "paper", "sandstone", "simplex", "yeti"
# Calls: <Anonymous> ... do.call -> <Anony
```

### Assignment Overview

The New York Times web site provides a rich set of APIs, as described [here](https://developer.nytimes.com/apis).  
You’ll need to start by signing up for an API key. Your task is to choose one of the New York Times APIs, construct an interface in R to read in the JSON data, and transform it into an R DataFrame.  

### Packages
```{r}
library(httr)
library(jsonlite)
library(reactable)
```

### Load API Data
After signing up for the API Key, I used the best selling nonfiction books reported by The New York Times for the week of 01/05/2020 which can also be verified on [The New York Times](https://www.nytimes.com/books/best-sellers/2020/01/05/combined-print-and-e-book-nonfiction/) website. 
Using the GET from the httr library, the API webpage was retrieved with he use of the API key requested. 
content function from the httr library was used along with the from JSON junction from the jsonlite library to extract the data into a list. 
```{r}
res = GET("https://api.nytimes.com/svc/books/v3/lists/2020-01-05/hardcover-nonfiction.json?api-key=AqosXChWd9icDpFtb4chdjBylJLrwsdY")
```

```{r}
api_data <- fromJSON(content(res, "text"))
#api_data is a list
typeof(api_data)
```

### Extracting and Transform
The next step was to extract the column data to create the desired dataframe using the base R functions cbind, as.data.frame, and colnames. Cbind was used to extract the columns and then as.data.frame was used to convert the columns into a dataframe. Finally, colnames was used to update my column names. 
```{r}
column_data <- cbind(api_data$results$books$rank, api_data$results$books$title,
                     api_data$results$books$author,api_data$results$books$description,
                     api_data$results$books$publisher, api_data$results$books$primary_isbn13)
book_df <- as.data.frame(column_data)

colnames(book_df) <- c("Rank", "Title", "Author", "Description", "Publisher", "ISBN 13")
```

### The New York Times Best Sellers January 05, 2020
For my dataframe I decided to extract Rank, Title, Author, Description, Publisher, and the ISBN 13 number for each best seller for the week of January 5, 2020. Below is a table created using the reactable package for the list of the 15 best sellers. 
```{r}
reactable(book_df, bordered = TRUE, striped = TRUE, highlight = TRUE, 
          filterable = TRUE,  showPageSizeOptions = TRUE, 
          pageSizeOptions = c(5, 10, 15), defaultPageSize = 5)
```

